# Introduction to the data

Before tackling analytics problem, we start by introducing data to be analyzed in later chapters. 

## Customer Data for Clothing Company

Our first data set represents customers of a clothing company who sells products in physical stores and online. This data is typical of what one might get from a company's marketing data base (the data base will have more data than the one we show here). This data includes 1000 customers: 

1. Demography
    - `age`: age of the respondent
    - `gender`: male/female
    - `house`: 0/1 variable indicating if the customer owns a house or not

1. Sales in the past year
    - `store_exp`: expense in store
    - `online_exp`: expense online
    - `store_trans`: times of store purchase
    - `online_trans`: times of online purchase

1. Survey on product preference

It is common for companies to survey their customers and draw insights to guide future marketing activities. The survey is as below:

How strongly do you agree or disagree with the following statements:

1. Strong disagree
1. Disagree
1. Neither agree nor disagree
1. Agree
1. Strongly agree

- Q1. I like to buy clothes from different brands
- Q2. I buy almost all my clothes from some of my favorite brands
- Q3. I like to buy premium brands
- Q4. Quality is the most important factor in my purchasing decision
- Q5. Style is the most important factor in my purchasing decision
- Q6. I prefer to buy clothes in store
- Q7. I prefer to buy clothes online
- Q8. Price is important 
- Q9. I like to try different styles
- Q10. I like to make decision myself and don't need too much of others' suggestions 

There are 4 segments of customers: 

1. Price
1. Conspicuous
1. Quality
1. Style

Let's check it:

```{r, echo=FALSE}
sim.dat <- read.csv("http://bit.ly/2P5gTw4")
```

```{r}
str(sim.dat,vec.len=3)
```

Refer to Appendix for the simulation code.

## Customer Satisfaction Survey Data from Airline Company

This data set is from a customer satisfaction survey for three airline companies. There are `N=1000` respondents and 15 questions. The market researcher asked respondents to recall the experience with different airline companies and assign a score (1-9) to each airline company for all the 15 questions. The higher the score, the more satisfied the customer to the specific item. The 15 questions are of 4 types (the variable names are in the  parentheses):


- How satisfied are you with our______?

1. Ticketing
    - Ease of making reservation（Easy_Reservation）
    - Availability of preferred seats（Preferred_Seats）
    - Variety of flight options（Flight_Options）
    - Ticket prices（Ticket_Prices）
1. Aircraft
    - Seat comfort（Seat_Comfort）
    - Roominess of seat area（Seat_Roominess）
    - Availability of Overhead（Overhead_Storage）
    - Cleanliness of aircraft（Clean_Aircraft）
1. Service
    - Courtesy of flight attendant（Courtesy）
    - Friendliness（Friendliness）
    - Helpfulness（Helpfulness）
    - Food and drinks（Service）
1. General
    - Overall satisfaction（Satisfaction）
    - Purchase again（Fly_Again）
    - Willingness to recommend（Recommend）

Now check the data frame we have:

```{r, echo=FALSE}
rating<-read.csv("http://bit.ly/2TNQ6TK")
```

```{r}
str(rating,vec.len=3)
```

Refer to Appendix for the simulation code.

## Swine Disease Breakout Data

The swine disease data includes simulated 120 survey questions from 800 farms. There are three choices for each question. The outbreak status for the $i^{th}$ farm is generated from a $Bernoulli(1, p_i)$ distribution with $p_i$ being a function of the question answers: 

$$ln(\frac{p_i}{1-p_i})=\beta_0 + \Sigma_{g=1}^G\mathbf{x_{i,g}^T\beta_{g}}$$

where $\beta_0$ is the intercept, $\mathbf{x_{i,g}}$ is a three-dimensional indication vector for question answer and $\mathbf(\beta_g)$ is the parameter vector corresponding to the $g^{th}$ predictor. Three types of questions are considered regarding their effects on the outcome. The first forty survey questions are important questions such that the coefficients of the three answers to these
questions are all different:

$$\mathbf{\beta_g}=(1,0,-1)\times \gamma,\ g=1,\dots,40$$

The second forty survey questions are also important questions but only one answer has a coefficient that is different from the other two answers:

$$\mathbf{\beta_g}=(1,0,0)\times \gamma,\ g=41,\dots,80$$

The last forty survey questions are also unimportant questions such that all three answers have the same coefficients:

$$\mathbf{\beta_g}=(0,0,0)\times \gamma,\ g=81,\dots,120$$

The baseline coefficient $\beta_0$ is set to be $-\frac{40}{3}\gamma$ so that on average a farm have 50% of chance to have an outbreak. The parameter $\gamma$ in the above simulation is set to control the strength of the questions' effect on the outcome. In this simulation study, we consider the situations where $\gamma = 0.1, 0.25, 0.5, 1, 2$. So the parameter settings are:

$$\mathbf{\beta^{T}}=\left(\underset{question\ 1}{\frac{40}{3},\underbrace{1,0,-1}},...,\underset{question\ 40}{\underbrace{1,0,-1}},\underset{question\ 41}{\underbrace{1,0,0}},...,\underset{question\ 80}{\underbrace{1,0,0}},\underset{question\ 81}{\underbrace{0,0,0}},...,\underset{question\ 120}{\underbrace{0,0,0}}\right)*\gamma$$

For each value of $\gamma$, 20 data sets are simulated. The bigger $\gamma$ is, the larger the corresponding parameter. We provided the data sets with $\gamma = 2$. Let's check the data:

```{r}
disease_dat<-read.csv("http://bit.ly/2KXb1Qi")
# only show the last 7 columns here
head(subset(disease_dat,select=c("Q118.A","Q118.B","Q119.A",
                                 "Q119.B","Q120.A","Q120.B","y"))) 
```

Here `y` indicates the outbreak situation of the farms. `y=1` means there is an outbreak in 5 years after the survey. The rest columns indicate survey responses. For example `Q120.A = 1` means the respondent chose `A` in Q120. We consider `C` as the baseline. 

Refer to Appendix for the simulation code.

## MNIST Dataset

## IMDB Dataset
