# Measuring Performance

To compare different models, we need a way to measure model performance. There are various metrics to use. To better understand the strengths and weaknesses of a model, you need to look at it through multiple metrics. In this chapter, we will introduce some of the most common performance measurement metrics.

## Regression Model Performance

**MSE and RMSE**

Mean Squared Error (MSE) measures the average of the squares of the errors—that is, the average squared difference between the estimated values and the actual value. The Root Mean Squared Error (RMSE) is the root square of the MSE. 

$$MSE=\frac{1}{n}\sum_{i=1}^{n}(y_{i}-\hat{y}*{i})^{2}$$
$$RMSE=\sqrt{\frac{1}{n}\sum*{i=1}^{n}(y_{i}-\hat{y}_{i})^{2}}$$

Both are the common measurements for the regression model performance.  Let's use the previous `income` prediction as an example. Fit a simple linear model:

```{r}
sim.dat <- read.csv("http://bit.ly/2P5gTw4")
fit<- lm(formula = income ~ store_exp + online_exp + store_trans + 
    online_trans, data = sim.dat)
summary(fit)
```

The fitted results `fit` shows the RMSE is `r as.integer(round(summary(fit)$sigma,0))` (at the bottom of the output after `Residual standard error:`).

Another common performance measure for the regression model is R-Squared, often denoted as $R^2$. It is the square of the correlation between the fitted value and the observed value. It is often explained as the percentage of the information in the data that can be explained by the model. The above model returns a R-squared＝`r round(summary(fit)$adj.r.squared,2)`, which indicates the model can explain `r round((summary(fit)$adj.r.squared)*100,0)`% of the variance in variable `income`.  While $R^2$ is easy to explain, it is not a direct measure of model accuracy but correlation. Here the $R^2$ value is not low but the RMSE is  `r round(summary(fit)$adj.r.squared,2)`  which means the average difference between model fitting and the observation is  `r round(summary(fit)$sigma,0)`. It is a big discrepancy from an application point of view. When the response variable has a large scale and high variance, a high $R^2$ doesn't mean the model has enough accuracy. It is also important to remember that $R^2$ is dependent on the variation of the outcome variable. If the data has a response variable with a higher variance, the model based on it tends to have a higher $R^2$.

We used $R^2$ to show the impact of the error from independent and response variables in Chapter \@ref(modeltuningstrategy) where we didn't consider the impact of the number of parameters (because the number of parameters is very small compared to the number of observations).  However, $R^2$ increases as the number of parameters increases. So people usually use Adjusted R-squared which is designed to mitigate the issue. The original $R^2$ is defined as: 

$$R^{2}=1-\frac{RSS}{TSS}$$

where $RSS=\sum_{i=1}^{n}(y_{i}-\hat{y_{i}})^{2}$ and $TSS=\sum_{i=1}^{n}(y_{i}-\bar{y})^{2}$.

Since RSS is always decreasing as the number of parameters increases, $R^2$ increases as a result.  For a model with $p$ parameters, the adjusted $R^2$ is defined as:

$$Adjusted\ R^{2}=1-\frac{RSS/(n-p-1)}{TSS/(n-1)}$$

To maximize the adjusted $R^{2}$ is identical to minimize $RSS/(n-p-1)$. Since the number of parameters $p$ is reflected in the equation, $RSS/(n-p-1)$ can increase or decrease as $p$ increases. The idea behind this is that the adjusted R-squared increases if the new variable improves the model more than would be expected by chance. It decreases when a predictor improves the model by less than expected by chance. While values are usually positive, they can be negative as well. 

Another measurement is $C_{p}$. For a least squared model with $p$ parameters:

$$C_{p}=\frac{1}{n}(RSS+2p\hat{\sigma}^{2})$$

where $\hat{\sigma}^{2}$ is the estimator of the model random effect $\epsilon$. $C_{p}$ is to add penalty $2p\hat{\sigma}^{2}$ to the training set $RSS$. The goal is to adjust the over-optimistic measurement based on training data. As the number of parameters increases, the penalty increases.  It counteracts the decrease of $RSS$ due to increasing the number of parameters. We choose the model with a smaller $C_{p}$.

Both AIC and BIC are based on the maximum likelihood. In linear regression, the maximum likelihood estimate is the least squared estimate. The definitions of the two are:

$$AIC=n+nlog(2\pi)+nlog(RSS/n)+2(p+1)$$

$$BIC=n+nlog(2\pi)+nlog(RSS/n)+log(n)(p+1)$$

R function AIC() and BIC() will calculate the AIC and BIC value according to the above equations. Many textbooks ignore content item $n+nlog(2\pi)$, and use $p$ instead of $p+1$. Those slightly different versions give the same results since we are only interested in the relative value. Comparing to AIC, BIC puts a heavier penalty on the number of parameters. 

## Classification Model Performance

This section focuses on performance measurement for models with a categorical response. The metrics in the previous section are for models with a continuous response and they are not appropriate in the context of classification. Most of the classification problems are dichotomous, such as an outbreak of disease, spam email, etc. There are also cases with more than two categories as the segments in the clothing company data. We use swine disease data to illustrate different metrics. Let's train a random forest model as an example. We will discuss the model in Chapter \@ref(treemodel).

```{r}
library(dplyr)
library(randomForest)
library(caret)
library(readr)
disease_dat <- read.csv("http://bit.ly/2KXb1Qi")
# you can check the data using glimpse()
# glimpse(disease_dat)
```

Separate the data to be training and testing sets.  Fit model using training data (`xTrain` and `yTrain`) and applied the trained model on testing data (`xTest` and `yTest`) to evaluate model performance. We use 70% of the sample as training and the rest 30% as testing.

```{r}
# separate the data to be training and testing
trainIndex <- createDataPartition(disease_dat$y, p = 0.8, 
    list = F, times = 1)
xTrain <- disease_dat[trainIndex, ] %>% dplyr::select(-y)
xTest <- disease_dat[-trainIndex, ] %>% dplyr::select(-y)
# the response variable need to be factor
yTrain <- disease_dat$y[trainIndex] %>% as.factor()
yTest <- disease_dat$y[-trainIndex] %>% as.factor()
```

Train a random forest model: 

```r
train_rf <- randomForest(yTrain ~ ., data = xTrain, mtry = trunc(sqrt(ncol(xTrain) - 
    1)), ntree = 1000, importance = T)
```

```{r, echo=FALSE}
load("../../GitHub/DataScientistR/Data/train_rf.RData")
```

Apply the trained random forest model to the testing data to get two types of predictions:

- probability (a value between 0 to 1)

```{r}
yhatprob <- predict(train_rf, xTest, "prob")
set.seed(100)
car::some(yhatprob)
```

- category prediction (0 or 1)

```{r}
yhat <- predict(train_rf, xTest)
car::some(yhat)
```

We will use the above two types of predictions to show different performance metrics.

**Kappa Statistics**

**Confusion Matrix** is a counting table to describe the performance of a classification model. For the true response `yTest` and prediction `yhat`, the confusion matrix is:

```{r}
yhat = as.factor(yhat) %>% relevel("1")
yTest = as.factor(yTest) %>% relevel("1")
table(yhat,yTest)
```

The top-left and bottom-right are the numbers of correctly classified samples. The top-right and bottom-left are the numbers of wrongly classified samples.  A general confusion matrix for a binary classifier is following: 

|   | Predicted Yes  | Predicted No | 
|:---:|:---:|:---:|
| Actual Yes  |  TP | FN  |
| Actual No  | FP  |  TN | 

where  TP is true positive, FP is false positive, TN is true negative, FN is false negative. The cells along the diagonal line from top-left to bottom-right contain the counts of correctly classified samples. The cells along the other diagonal line contain the counts of wrongly classified samples. The most straightforward performance measure is the **total accuracy** which is the percentage of correctly classified samples: 

$$Total\ accuracy = \frac{TP+TN}{TP+TN+FP+FN}$$

You can calculate the total accuracy when there are more than two categories. This statistic is straightforward but has some disadvantages. First, it doesn't differentiate different error types. In a real application, different types of error may have different impacts. For example, it is much worse to tag an important email as spam and miss it than failing to filter out a spam email.  Provost et al. [@Provost1998] discussed in detail the problem of using total accuracy on different classifiers. There are some other metrics based on the confusion matrix that measure different types of error. P**recision** is a metric to measure how accurate positive predictions are (i.e. among those emails predicted as spam, how many percentages of them are spam emails?):

$$precision = \frac{TP}{TP+FP}$$

**Recall sensitivity** is to measure the coverage of actual positive samples (i.e. among those spam emails, how many percentages of them are predicted as spam) :

$$Recall\ sensitivity = \frac{TP}{TP+FN}$$

Specificity is to measure the coverage of actual negative samples (i.e. among those non-spam emails, how many percentages of them pass the filter):

$$Specificity = \frac{TN}{TN+FP}$$

Since wrongly tagging an important email as spam has a bigger impact, in the spam email case, we want to make sure the model specificity is high enough.

Second, total accuracy doesn't reflect the natural frequencies of each class. For example, the percentage of fraud cases for insurance may be very low, like 0.1%. A model can achieve nearly perfect accuracy (99.9%) by predicting all samples to be negative. The percentage of the largest class in the training set is also called the no-information rate. In this example, the no-information rate is 99.9%. You need to get a model that at least beats this rate. 

Another metric is the Kappa statistic. It measures the agreement between the observed and predicted classes. It was originally come up by Cohen etc. [@Cohen1960]. Kappa takes into account the accuracy generated simply by chance. It is defined as:

$$Kappa=\frac{P_{0}-P_{e}}{1-P_{e}}$$

Let $n=TP+TN+FP+FN$ be the total number of samples, where $P_{0}=\frac{TP+TN}{n}$ is the observed accuracy, $P_{e}=\frac{(TP+FP)(TP+FN)+(FN+TN)(FP+TN)}{n^{2}}$ is the expected accuracy based on the marginal totals of the confusion matrix.  Kappa can take on a value from  -1 to 1. The higher the value,  the higher the agreement. A value of 0 means there is no agreement between the observed and predicted classes, while a value of 1 indicates perfect agreement. A negative value indicates that the prediction is in the opposite direction of the observed value. The following table may help you "visualize" the interpretation of kappa [@landis1977]:

| Kappa | Agreement |
|:-----:|:-----:|
| < 0 | Less than chance agreement |
| 0.01–0.20 | Slight agreement |
| 0.21– 0.40 | Fair agreement |
| 0.41–0.60 | Moderate agreement |
| 0.61–0.80 | Substantial agreement | 
| 0.81–0.99 | Almost perfect agreement |

In general, a value between 0.3 to 0.5 indicates a reasonable agreement. If a model has a high accuracy of 90%, while the expected accuracy is also high, say 85%. The Kappa statistics is $\frac{1}{3}$.  It means the prediction and the observation have a fair agreement. You can calculate Kappa when the number of categories is larger than 2. The package `fmsb` has a function  `Kappa.test()` to calculate Cohen's Kappa statistics. The function can also return the hypothesis test result and a confidence interval. Use the above observation vector `yTest` and prediction vector `yhat` as an example, you can calculate the statistics:

```{r}
# install.packages("fmsb")
kt<-fmsb::Kappa.test(table(yhat,yTest))
kt$Result$estimate
```

